% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}

    This PhD thesis is a product of the support of my supervisors, advisors, collaborators, family, and friends, and would not have been possible without all their encouragement and support.

    First and foremost, I am incredibly grateful to my supervisors Dr Anne Presanis, Dr Paul Birrell, and Professor Daniela De Angelis who have challenged, supported, and inspired me in equal measure throughout four extraordinary years. I would like to especially thank Anne whose thoughtful guidance and encouragement has shaped me as a researcher.

    I am also indebted to my advisors and collaborators, both at the MRC Biostatistics Unit and further afield. Particular thanks to Dr Shaun Seaman, Dr Chris Jackson, Dr Brian Tom, and Dr Andre Charlett for all the advice and useful discussion; to Dr Sara Croxford, Meaghan Kall, Dr Alison Brown, Cuong Chau, and Dr Valerie Delpech for encouraging me to begin this journey and for their support ever since; and to Dr Victoria Hall, Sarah Foulkes, Jameel Khawam, Dr David Wyllie, Ranya Mulchandani, Dr Sema Mandal, Suzanne Elgohari, and Dr Ferdinando Insalata for stimulating conversation and collaboration during the height of the pandemic.
    
    Thank you to the UK Medical Research Council for supporting my research, to Dr Brian Tom, Dr Rob Goudie, and Dr Angela Talbot for their support of the PhD programme, and to my college tutors Dr Lori Passmore and Dr Holly Hedgeland for pastoral support.

    Thank you to my dad, brother, and all of my family for their genuine interest and support over the years. Thank you to my oldest friends Jez, Chris, Ollie, Rafal, and Gorkem, who I could always rely on to not take things too seriously. Thank you to all my friends in Cambridge, especially to Debbie, Anna, Hubert, Simone, Esther, Sarah, Lukas, Ruta, Nia, and Ciara. And to Katie, Krathika, Tori, and Guilia thank you for making our house a home. Thank you to all my friends at the MRC Biostatistics Unit, especially Josh, Chiara, and Andrew who shared the ups and downs of PhD-ing during a pandemic. Thank you to all my friends at Clare Hall and CHBC for always keeping life interesting, and to the Newnham Beavers and leaders for reminding me of the value of community.

    Last but not least, I would also like to acknowledge all those living with HIV, those who were admitted to hospital during the COVID-19 pandemic, and all the SIREN study participants who contributed their data, as well as the doctors, nurses, data reporters, and research staff without whom this research would not have been possible.

\end{acknowledgements}
