%!TEX root = ../thesis.tex
% ******************************* Thesis Appendix A ****************************
\chapter{PhD publications, pre-prints and conference presentations}\label{sec:publications}

\begin{spacing}{0.9}

    \begin{refsection}
        \nocite{Kirwan2024-hj, Kirwan2022-za, Kirwan2022-ka, Kirwan2021-yl}
        \printbibliography[title={First-author peer-reviewed publications}, heading=subbibliography]
    \end{refsection}

    \begin{refsection}
        \nocite{Brizzi2019-yj, Hall2021-rh, Mulchandani2020-en, Allen2021-jo, Brizzi2021-zl, Docherty2021-es, Presanis2021-pv, Pantazis2021-dv, Brown2022-mn, Hall2022-ep, Wallace2022-vs, Jackson2022-lt, Hall2024-ai}
        \printbibliography[title={Contributing author peer-reviewed publications}, heading=subbibliography]
    \end{refsection}

    \begin{refsection}
        \nocite{Kirwan2021-sw}
        \printbibliography[title={First-author pre-prints}, heading=subbibliography]
    \end{refsection}

    \section*{Conference presentations and posters}

    \begin{description}
        \item Sensitivity of an HIV back-calculation model to counterfactual COVID-19 lockdown assumptions, \textbf{Greek Stochastics 2022}, Aug 2022 [Oral presentation], \textbf{UKHSA 2022}, Oct 2022 [Poster presentation], and \textbf{MRC Biostatistics Unit Armitage week 2023} [Poster presentation]. % add ECDC meeting
        \item Protection of SARS-CoV-2 vaccination and prior infection against future infection. \textbf{MRC Biostatistics Unit Armitage week 2022}, Sep 2022 [Oral presentation], and \textbf{RSS 2022}, Sep 2022 [Poster presentation].
        \item Trends in COVID-19 hospital outcomes in England before and after vaccine introduction, 2020--2021: a cohort study. \textbf{UKHSA 2022}, Oct 2022 [Oral presentation].
        \item Protection of COVID-19 booster vaccination and prior infection against symptomatic and asymptomatic SARS-CoV-2. \textbf{UKHSA 2023}, Nov 2023 [Poster presentation] and \textbf{ESCAIDE 2023}, Nov 2023 [Oral presentation by co-author].
        \item Monitoring protection from SARS-CoV-2 infection and COVID-19 vaccination in the UK SIREN healthcare worker cohort 2020--2024: Assessing the impact of vaccine boosters. \textbf{ECCMID 2024}, Apr 2024 [Poster presentation].
    \end{description}

    \section*{Other academic activities}

    \begin{description}
        \item MRC Biostatistics Unit PhD student representative and organised the annual PhD student Symposium, 2021--2022.
        \item Invited speaker at Scientific Pandemic Influenza Group on Modelling (SPI-M) sub-group meeting, Feb 2022.
        \item Interviewer for Armitage week:\ In conversation with\ldots\ Prof.\ Nicholas Jewell, Sep 2022. \url{https://www.mrc-bsu.cam.ac.uk/event/19th-armitage-workshop-and-lecture/}
        \item Course demonstrator on Bayesian Modelling short course delivered to PhD students in Addis Ababa, Ethiopia, Oct 2022. \url{https://www.mrc-bsu.cam.ac.uk/blog/bsu-researchers-teach-bayesian-statistics-course-in-ethiopia/}
        \item Recipient of SPI-M Award for Modelling and Data Support (SAMDS), Oct 2022. \url{https://www.mrc-bsu.cam.ac.uk/blog/mrc-biostatistics-unit-covid-19-working-group-receive-spi-m-o-award/}
        \item Invited speaker and panellist on European Testing Week/EuroTEST `Updated definition of HIV late diagnosis' webinar, Nov 2022. \url{https://www.testingweek.eu/resources/webinars-videos/etw-2022-updated-definition-of-hiv-late-diagnosis/}
        \item Course demonstrator on University of Cambridge Population Health Science MPhil, Bayesian Modelling and Infectious Disease Modelling modules, Easter 2022 and 2023.
        \item Invited speaker at Joint Committee on Vaccination and Immunisation (JCVI) COVID-19 meeting, May 2023.
    \end{description}

\end{spacing}
